---
title: "Poverty and Inequality Platform Methodology Handbook"
zotero: "PovcalNet Team"
url: "https://povcalnet-team.github.io/Methodology"
github-repo: PovcalNet-Team/Methodology
site: "bookdown::bookdown_site"
output:
  bookdown::gitbook: default
documentclass: book 
editor_options: 
  markdown: 
    wrap: 72
bibliography: references.bib
link-citations: yes
links-as-notes: true
colorlinks: yes
---

```{=html}
<style type="text/css">
h1 { /* Header 1 */
  font-size: 24px;
}
h2 { /* Header 2 */
    font-size: 18px;
}
h3 { /* Header 3 */
    font-size: 14px;
}
body {
text-align: justify}

</style>
```
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE) 
```

# Non-technical summary {-}

This site describes how global and regional poverty rates are calculated. This particular page summarizes the methodology in non-technical terms using four steps. The summary is followed by four chapters that dig into these steps in detail.

The four steps are illustrated in Figure 1 and outlined below in more detail. First, household survey data are obtained from relevant sources. Second, the survey data are used to create an estimate of households' income or consumption, known as welfare aggregates. Third, the welfare aggregates are adjusted for differences in price levels across countries and over time to foster international comparability. Upon this, poverty can be estimated for a particular country for a particular year. Fourth, the estimates of poverty are extrapolated or interpolated to a common year and the population-weighted poverty rate is calculated.

```{r, echo=FALSE}
DiagrammeR::grViz("digraph {
  graph [layout = dot, rankdir = TD]
  
  node [shape = box,style=filled, color=lightblue2]        
  box1 [label = 'Step 1\nAcquiring household survey data']
  box2 [label = 'Step 2\nConstructing welfare aggregates']
  box3 [label = 'Step 3\nGenerating survey estimates of poverty and inequality']
  box4 [label = 'Step 4\nGenerating global and regional estimates of poverty']
  box1 -> box2 -> box3 -> box4
  }", 
                  height = 500)
```

## Step 1: [Acquiring household survey data](#acquiring) {.unnumbered}

Poverty rates are estimated from [selected](#selection) household surveys. For the most part, these are surveys that ask a [representative](#representativeness) subset of households in a country about their consumption or income. Most household surveys from developing countries [are obtained](#selection) through collaboration with countries' National Statistical Offices, while most data for high-income countries are obtained from the [EU Statistics on Income and Living Conditions (EU-SILC)](%5Bhttps://ec.europa.eu/eurostat/web/microdata/european-union-statistics-on-income-and-living-conditions) or from the [Luxembourg Income Study (LIS) Database](https://www.lisdatacenter.org). At times, household-level data cannot be obtained in which case data [aggregated](#grouped) at a higher level are obtained.

## Step 2: [Constructing welfare aggregates](#welfareaggregate) {.unnumbered}

The Poverty and Inequality Platform utilizes a monetary measure of poverty. Monetary poverty is estimated from an aggregation of household's income or from the monetary value of their consumption. We refer to such aggregates jointly as welfare aggregates. Welfare aggregates are harmonized across countries and over time to maximize [comparability](#comparability), but decisions such as [whether income or consumption is used](#incomeorconsumption), [the design of the questionnaire](#questionnaire) of the household survey, what components that are included in the welfare aggregate, and whether [price differences within a country](#spatialtemporal) are accounted for, means that full comparability is not feasible. 

## Step 3: [Generating survey estimates of poverty and inequality](#surveyestimates) {.unnumbered}

Welfare aggregates are expressed in local currencies in the prices prevailing at the time of the collection of the data. To compare the consumption of an Indian household in 2011 to the consumption of a Nigerian household in 2018, the welfare aggregate needs to be expressed in the same prices. To this end, [consumer price indices](#cpis) are used to express all welfare aggregates in local 2011 prices, and [purchasing power parities](#ppps) are used to account for price differences between countries. Once all welfare aggregates are expressed in the same units, they can be used to estimate extreme poverty. Extreme poverty is measured as the share of the population of a country whose welfare per day falls short of the international poverty line. [The international poverty line](#IPL) is the typical poverty line used by the poorest countries of the world.

## Step 4: [Generating global and regional estimates of poverty](#lineupestimates) {.unnumbered}

Most countries do not conduct household surveys every year. Yet, to estimate regional and global poverty for a particular year, one needs an estimate of poverty for every country for the year in question. To this end, the estimates of poverty from surveys are [extrapolated](#extrapolations) or [interpolated](#interpolations). The extrapolations assume that everyone's income or consumption grows in accordance with growth rates from [national accounts](#nationalaccounts) between the time of the survey and the year in question. For [countries without any data](#nodata) at all, it is assumed that their poverty rate is equal to the average poverty rate in their region. To avoid presenting regional and global numbers with insufficient recent survey data, [coverage rules](#coveragerule) are used to determine whether a particular reference year has sufficient surveys data close to the year.
