---
title: "Poverty and Inequality Platform Methodology Handbook"
zotero: "PovcalNet Team"
url: "https://povcalnet-team.github.io/Methodology"
github-repo: PovcalNet-Team/Methodology
site: "bookdown::bookdown_site"
output:
  bookdown::gitbook:
    config:
      search: yes
documentclass: book 
editor_options: 
  markdown: 
    wrap: 72
bibliography: references.bib
biblio-style: "apalike"
link-citations: yes
links-as-notes: true
colorlinks: yes
---

```{=html}
<style type="text/css">
h1 { /* Header 1 */
  font-size: 24px;
}
h2 { /* Header 2 */
    font-size: 18px;
}
h3 { /* Header 3 */
    font-size: 14px;
}
body {
text-align: justify}

</style>
```
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE) 
```

*Suggested citation: World Bank. 2022. "Poverty and Inequality Platform
Methodology Handbook." Edition 2022-04. Available at
[<https://worldbank.github.io/PIP-Methodology-2022-04>](https://worldbank.github.io/PIP-Methodology/index.html).*

# Non-technical summary {.unnumbered}

This site describes how global and regional poverty rates are
calculated. This particular page summarizes the methodology in
non-technical terms using five steps. The summary is followed by five
chapters that dig into these steps in detail.

The five steps are illustrated in the figure below and summarized after
the figure. First, household survey data are obtained from relevant
sources. Second, the survey data are used to create an estimate of
households' income or consumption, known as welfare aggregates. Third,
the welfare aggregates are adjusted for differences in price levels
across countries and over time to foster international comparability.
Fourth, poverty and inequality are estimated for a particular country
for a particular year. Fifth, the estimates of poverty are extrapolated
or interpolated to a common year and the population-weighted poverty
rate is calculated.

All decisions, assumptions, and protocols involved in these steps and in
the chapters that follow are governed by the World Bank's Global Poverty
Working Group, which is composed of staff from the Poverty and Equity
Global Practice, the Development Data Group, and the Development
Research Group.

```{r, echo=FALSE}
DiagrammeR::grViz("digraph {
  graph [layout = dot, rankdir = TD]
  
  node [shape = box,style=filled, color=lightblue2]        
  box1 [label = 'Step 1\nAcquiring household survey data']
  box2 [label = 'Step 2\nConstructing welfare aggregates']
  box3 [label = 'Step 3\nConverting welfare aggregates']
  box4 [label = 'Step 4\nCalculating survey estimates of poverty and inequality']
  box5 [label = 'Step 5\nCalculating global and regional poverty']
  box1 -> box2 -> box3 -> box4 -> box5
  }", 
                  height = 500)
```

## Step 1: [Acquiring household survey data](#acquiring) {.unnumbered}

Poverty rates are estimated from [selected](#selection) household
surveys. In general, the surveys used ask a
[representative](#representativeness) subset of households in a country
about their consumption or income. These surveys are often the official
surveys used by countries to monitor and report on poverty. Most
household surveys from developing countries [are obtained](#selection)
through collaboration with countries' National Statistical Offices,
while most data for high-income countries are obtained from the [EU
Statistics on Income and Living Conditions
(EU-SILC)](https://ec.europa.eu/eurostat/web/microdata/european-union-statistics-on-income-and-living-conditions)
or from the [Luxembourg Income Study (LIS)
Database](https://www.lisdatacenter.org). At times, household-level data
cannot be obtained in which case, as a second best,
[aggregated](#grouped) data are used, such as income or consumption by
population deciles.

## Step 2: [Constructing welfare aggregates](#welfareaggregate) {.unnumbered}

The Poverty and Inequality Platform (PIP) primarily utilizes a monetary
measure of poverty. Monetary poverty is estimated from an aggregation of
household's income or from the monetary value of their consumption. We
refer to such aggregates jointly as welfare aggregates. Welfare
aggregates are harmonized across countries and over time to maximize
[comparability](#comparability), but country-specific decisions on
issues such as [whether income or consumption is
used](#incomeorconsumption), [the design of the
questionnaire](#questionnaire) of the household survey, what [components
that are included](#imputedrent) in the welfare aggregate, and whether
[price differences within a country](#spatialtemporal) are accounted
for, imply that full comparability is not feasible.

## Step 3: [Converting welfare aggregates](#convert) {.unnumbered}

Welfare aggregates are often expressed in local currencies in the prices
prevailing at the time of the collection of the data. To compare the
consumption of an Indian household in 2011 to the consumption of a
Nigerian household in 2018, the welfare aggregates need to be expressed
in the same prices. To this end, [consumer price indices](#CPIs) are
used to express all welfare aggregates in local 2017 prices, and
[purchasing power parities](#PPPs) are used to account for price
differences between countries. Once all welfare aggregates are expressed
in the same units, a common poverty line is needed to estimate poverty.
[The international poverty line](#IPL) used in PIP is constructed from
the poverty lines used by the poorest countries of the world.

## Step 4: [Calculating survey estimates of poverty and inequality](#surveyestimates) {.unnumbered}

Armed with welfare distributions expressed in 2017 PPPs and an
international poverty line, poverty and inequality can be calculated and
compared across countries and over time. PIP contains a range of
different [monetary poverty measures](#monetarypoverty), a
[multidimensional poverty measure](#mpm), [inequality
measures](#inequality), and [other distributional statistics](#dist).

## Step 5: [Calculating global and regional poverty](#lineupestimates) {.unnumbered}

Most countries do not conduct household surveys every year. Yet, to
estimate regional and global poverty for a particular year, one needs an
estimate of poverty for every country for the year in question. When a
poverty estimate is not available for a given year, the estimates are
[extrapolated](#extrapolations) or [interpolated](#interpolations) from
other years. The extrapolations assume that everyone's income or
consumption grows in accordance with per capita growth rates from
[national accounts](#nationalaccounts) between the time of the survey
and the year in question. For [countries without any household
data](#nodata) at all, it is assumed that their poverty rate is equal to
the population-weighted average poverty rate in their
[region](#regionsandcountries). To ensure the quality of the regional
and global numbers, [coverage rules](#coverage) are used to determine
whether a particular reference year has sufficient nearby survey data
for global and regional numbers to be presented.
