---
title: "Poverty and Inequality Platform Methodology"
output:
  html_document:
    number_sections: yes
  word_document: default
  pdf_document: default
bibliography: references.bib
---

```{=html}
<style type="text/css">
h1 { /* Header 1 */
  font-size: 28px;
}
h2 { /* Header 2 */
    font-size: 22px;
}
</style>
```
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

[Daniel]{style="color:red"}

This page explains the methods used to derive the global, region, and country
level poverty counts. The page is divided into six sections:

*1. Acquiring household survey data:*

*2. Constructing welfare aggregates:*

*3. Generating survey estimates:*

*4. Calculating poverty and inequality:*

*5. Generating global and regional estimates:*

*6. Refernences:*

Please cite this page as:

World Bank. 2020. "Poverty and Inequality Platform Methodology." March 2020
version. Available at [link to follow].

For old versions of this page, please see [link to follow].

# Acquiring household survey data

[Daniel]{style="color:red"}

## Selection criteria and source of data

[Daniel]{style="color:red"}

For developing countries, mostly 0fficial surveys used by the government for
monitoring poverty. Occasionally surveys are dropped due to quality concerns
with the survey design, welfare aggregate or auxilliary data. Decision made by
GPWG, reason summarized in What's New.

The data mostly come from household surveys, explain very briefly what a
household survey is, that NSOs conduct them. EU-SILC and LIS for some countries,
a subset of these based on admin data. Link to more country information (country
methodology pages).

## Representativeness

[Daniel]{style="color:red"}

Discuss within-country spatial coverage. Mention that non-national surveys are
also available but not used for regional/global numbers (Argentina exception).
Temporal coverage; often covers one year, \#\# Binned vs. micro data
[Daniel]{style="color:red"}

Mention countries exclusevely relying on binned data.

## How data are obtained

[Daniel]{style="color:red"}

Mostly receive data through country engagement, poverty economist, processed and
harmonized through GMD.

# Constructing welfare aggregates

[Daniel]{style="color:red"}

## Income or consumption

[Marta]{style="color:red"}

Explain briefly each, advantages and disadvantages of each (PSPR 2018 annex has
material on this). Explain how zero/negative incomes are treated.

## Within-survey spatial/temporal deflation

[Samuel]{style="color:red"}

Explain that some countries use spatial/temporal deflation but not all. Explain
what spatial temporal deflation means. Mention India/China/Indonesia. Explain
decimal years and that for EU-SILC/LIS, the reported year is the year before the
survey year.

## Questionnaire design

[Daniel]{style="color:red"}

Discuss the relevance of questionnaire design for the consumption aggregate
(perhaps refer to some of the papers on the topic). Use the India MMRP/URP
example (PSPR 2018 has material).

## Imputed rent

During the process of constructing a welfare aggregate, housing is arguably the
most important and difficult of all the components to include. From a conceptual
point of view, housing could be understood as any other durable good. Its
present value is not relevant for the analysis of current welfare because
purchasing a house is such a large and relatively rare expenditure that is
should not be included the in welfare aggregate.[@deaton2002, p.35]. The
objective is instead on measuring the value of the of occupying the dwelling for
the length of the survey period.

Whether the welfare aggregate is based on consumption or income, accounting for
housing is equally important. When the welfare aggregate is based on
consumption, household surveys usually record the value of rent paid by market
tenants. Since rent is the market value of occupying a house for a given period
of time, it is feasible and empirically viable to estimate the flow of housing
services. Yet, the same information is not available for household owners, even
though they also "consume" housing services. By not including the value of
housing "consumed" by owners, it would look like two household with the very
same consumption patterns have different welfare just because one pays rent and
the other does not. In fact, the market tenant would look better off than the
household owner because the consumption of the former is higher than the one of
the latter.

When the welfare aggregate is based on income the logic is similar. In this
case, the welfare aggregate should be composed by the remuneration of all the
assets of the household including labor, capital, and durable goods such as
houses and cars [@gasparini2004]. Families that own their housing receive an
implicit value that is equivalent to the amount of money they would have to pay
in the market if they had to rent a dwelling similar to the one they are
currently living in.

In the context of the Global Monitoring Database (GMD), each country has its own
methodological approach to input the implicit value of housing to dwelling
owners. The Latin American Region (LAC) is the only one that applies the same
methodology, or at least principles, to input rents in all its countries. All
the other regions do not have a consolidated approach so they adopt whatever the
National Statistics Office (NSO) of each country provides.

The methodology used by the LAC region could be found here [[Ask for document to
the LAC team]{style="color:red"}]. Yet, below you may find a short summary of
the decision tree used in LAC to account for housing rents.

1.  If the National Statistics Office (NSO) of the country releases their own
    estimate to account for own-housing rent, the SEDLAC project adopts that
    variable.  

```{=html}
<!-- -->
```
2.  If the NSO does not release a variable to account for own-housing rent, the
    value reported by owners in the household survey to the question "How much
    would you have to pay a month if you had to rent this house?" is taken as
    the implicit rent.  

```{=html}
<!-- -->
```
3.  If the questionnaire does not ask the self-reported rent question to owners,
    the utility obtained from living in a dwelling is obtained from a hedonic
    model that uses quantile regression of the sum of the utilities derived from
    the characteristics of the dwelling (e.g., geographical location,
    surrounding characteristics, and structural attributes), which is the
    implicit rent when monetarized [@rosen1974].

4.  Finally, if the questionnaire does not allow to identify ownership of the
    dwelling, no rent is imputed to owners. That is, the imputation is 0% of
    total income.

Step 3 in the list above. which is the most complex of all the steps in the
decision tree is applied only to four countries: Argentina, Brazil, Ecuador, and
Honduras.

According to Deaton and Zaidi [-@deaton2002], there are four main approaches to
input rents to dwelling owners plus a few other alternatives suggested in
Balcazar et al [-@balcazar2017]. Besides the LAC region, the rest of the
countries apply either one of these methodologies or perform no imputation at
all, which is equivalent to input zero (0) implicit rent to the dwelling owners.

[Daniel: I don't know the example of CHN. Do you have any documentation about
it? Is there something else you would like to add?]{style="color:blue"}

## Treatment of binned data

[Christoph and Samuel]{style="color:red"}

Let's put down the actual formulas we use here to convert binned data into
distributions.

## Equivalence scale

[Daniel]{style="color:red"}

Mention per capita assumptions, drawbacks of it, and why we are still assuming
it.

## Comparability database

In September 2019, the PovcalNet team and the Poverty GP put together the
[Comparability Over
Time](https://datacatalog.worldbank.org/dataset/comparability-over-time-country-level-international-poverty-measures)
dataset, which provides metadata on the comparability of poverty estimates from
[PovcalNet](http://iresearch.worldbank.org/PovcalNet "Opens in a new window")
within countries over time. The complete documentation of this dataset is
available in Atamanov et al [-@atamanov2019] (Section 4). The following excerpt
from this document clearly explains the motivation of this dataset:

> As countries frequently improve household surveys and measurement methodology,
> strict comparability of poverty estimates over time is often limited. Within a
> country, we assume comparability of poverty estimates over time unless there
> is a known change to survey methodology, measurement or data structure. Thus,
> the assessment of comparability is country-dependent and relies on the
> accumulation of knowledge from past and current Bank staff in the countries,
> as well as close dialogue with national data producers with knowledge of
> survey design and methodology

### Application {.unnumbered}

The comparability dataset is practically useful in any estimation that compares
welfare aggregates from household surveys over time. In the blog [*Apples to
apples --- PovcalNet introduces a new comparability
indicator*](https://blogs.worldbank.org/opendata/apples-apples-povcalnet-introduces-new-comparability-indicator)*,*
the authors present two examples on how to merge the comparability dataset with
the [PovcalNet](http://iresearch.worldbank.org/PovcalNet/home.aspx) estimates.
Both charts in the blog could be replicated in R and/or Stata by running the
scripts available in the
[worldbank/Global\_Poverty\_Blogs](https://github.com/worldbank/Global_Poverty_Blogs/tree/master/bg_povcalnet_comparability)
repository.

### Access the data {.unnumbered}

The dataset could be accessed in different ways. The official version is
available in the [Development Data
Hub](https://datacatalog.worldbank.org/dataset/comparability-over-time-country-level-international-poverty-measures)
of the World Bank. However, it is possible to access it directly from R, Stata,
or Python by calling this file,
"<https://development-data-hub-s3-public.s3.amazonaws.com/ddhfiles/506801/povcalnet_comparability.csv%22.>

#### R {.unnumbered}

```{r, eval = FALSE}
link <- "https://development-data-hub-s3-public.s3.amazonaws.com/ddhfiles/506801/povcalnet_comparability.csv"
df   <- readr::read_csv(link)
```

#### Stata {.unnumbered}

```{stata, eval = FALSE}
local dh "https://development-data-hub-s3-public.s3.amazonaws.com/ddhfiles/506801/povcalnet_comparability.csv"
import delimited using "`dh'", clear
```

#### Python {.unnumbered}

```{python, eval = FALSE}
import pandas as pd
link = "https://development-data-hub-s3-public.s3.amazonaws.com/ddhfiles/506801/povcalnet_comparability.csv"
df   = pd.read_csv(path)
```

Each observation of the comparability dataset represents a survey point which is
the combination of ISO3 country code (`countrycode`), `year`, survey acronym
(`survname`), Welfare type (i.e., income or consumption) (`datatype`), and
survey coverage (`coveragetype`). Variable `comparabiliy` specified the
comparability over time with the following logic. The oldest comparable series
in each country starts with the value zero (0) in the `comparability` variable.
When comparability is broken, the value changes to one (1) for the year of the
break and it goes on until the comparability is broken again in a subsequent
year. The process repeats until the most recent surveys point available. Thus,
to get the most recent comparable spell, you need the `max(comparability)` by
`countrycode, survname, datatype, coveragetype`.

# Generating survey estimates

[Daniel]{style="color:red"}

## Consumer Price Indices (CPIs)

[Samuel]{style="color:red"}

Short paragraph summary of why CPIs are needed, which CPIs we use, and link to
the CPI paper

## Purchasing Power Parities (PPPs)

[Samuel]{style="color:red"}

Short paragraph summary of why PPPs are needed, which PPPs we use and link to
the PPP paper

## Derivation of the international poverty line

[Samuel]{style="color:red"}

Explain in one paragraph or so how the international poverty line was derived
and link to the RCS and Ferreira et al. paper

## Derivation of other global poverty lines

[Samuel]{style="color:red"}

Explain how higher lines are derived and link to Jolliffe & Prydz paper

# Calculating poverty and inequality

[Daniel]{style="color:red"}

## FGT measures

[Aleksander and Samuel]{style="color:red"}

In this subsectoin and the one below, let's write the name of each measure in
Povcalnet and provide formulas for how they are derived. The FGT formulas should
be easily available online

## Inequality measures

[Aleksander and Samuel]{style="color:red"}.

Can we get to the source code find out how the various other measures are
calculated and list it here?

# Generating global and regional estimates

## National accounts data

[Nishant]{style="color:red"}

Explain in a short paragraph the national accounts data used, countries where we
use specific sources, and link to national accounts paper.

## Extrapolations

[Nishant]{style="color:red"}

Explain how/why we extrapolate, provide the formula, discuss the no-use of a
passthrough rate (perhaps refer to relevant literature on this) and link to
national accounts paper.

Explain the India exception.

## Interpolations

[Nishant]{style="color:red"}

Explain how/why we interpolate and provide the two different interpolation
formulas (in the national accounts paper).

## Choosing between consumption and income estimates

[Daniel]{style="color:red"}

## Regions and universe of countries

[Marta]{style="color:red"}

Explain the regions used and where the set of 218 countries comes from.

## Population

[Marta]{style="color:red"}

Explain the population data that we use.

## Missing countries

[Marta]{style="color:red"}

Explain how we treat countries with missing data and derive the regional/global
headcounts

## Coverage rule

[Marta]{style="color:red"}

Explain the coverage rule

# References
